---
title: "WrangleT3Downloads"
author: "Jean-Luc Jannink"
date: "2023-07-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries
```{r libraries}
library(tidyverse)
here::i_am("code/WrangleT3Downloads.Rmd")
```

## Simple function to impute missing data
```{r impute with glmnet}
# You enter the function with a matrix, individuals in rows, markers in columns
# It returns a matrix of the same dimensions but with no missing data
# I have only tested it with markers coded as -1, 0, 1 for AA, AB, and BB, but
# other codings would probably work too.
# Returns a matrix of the same dimensions, but with no missing data.
# Imputed values are real numbers (not integers). This may be problematic for
# some downstream mapping software.
# by default glmnet will look at ~100 different lambda penalty coefficients.
# It approximately doubles the speed to look at only 10 values. That probably
# lowers the accuracy by a couple percent, but not much more.
# Another thing that ~ doubles the speed is to do  5-fold rather than 10-fold cv
# The thing that makes the most difference is not putting ALL the other
# markers in as predictors, but only the top xx of them. I am using 60 now.
impute.glmnet <- function(matNA){
  require(glmnet)

  cvLambda <- exp(-(2:11))
  # Start with mean impute
  matNoNA <- apply(matNA, 2, 
                   function(v) {v[is.na(v)] <- mean(v, na.rm=TRUE); return(v)})
  # I am using 60 markers for prediction.
  # Make sure nPred less than number of markers in the matrix.
  nPred <- min(60, round(ncol(matNA) * 0.5))
  # Function to fill in one column with glmnet impute
  imputeMrk <- function(k){
    # Prevent imputations from going outside the original range
    varRange <- range(matNA[,k], na.rm=TRUE)
    isNA <- is.na(matNA[,k])
    # If the marker is monomorphic, impute with the sole value
    if (sd(matNA[,k], na.rm=TRUE) == 0){
      matNoNA[isNA,k] <<- matNA[which(!isNA)[1],k] 
    } else{
      corMrk <- abs(cov(matNA[,k], matNA, use="pairwise.complete.obs"))
      # Retain markers that correlate highly with marker to be imputed
      predMrk <- setdiff(order(corMrk,decreasing=TRUE)[1:nPred], k)
      cvModels <- cv.glmnet(x=matNoNA[!isNA,predMrk], 
                            y=matNA[!isNA,k], nfolds=5, lambda=cvLambda)
      pred <- predict(cvModels, s="lambda.min", 
                      newx=matNoNA[isNA,predMrk, drop=FALSE])
      pred[pred < varRange[1]] <- varRange[1]
      pred[pred > varRange[2]] <- varRange[2]
      matNoNA[isNA,k] <<- pred
    }
    return(k)
  }
  # Go in order from least to most missing (probably not really needed)
  sumIsNA <- apply(matNA, 2, function(v) sum(is.na(v)))
  imputeOrder <- order(sumIsNA)
  # Don't impute if none missing
  imputeOrder <- imputeOrder[sumIsNA[imputeOrder] > 0]
  dummy <- try(sapply(imputeOrder, imputeMrk), silent=TRUE)
  return(matNoNA)
}
```

## Read and wrangle the marker data  
After this `mrkDatT` is a tibble with accessions in rows and markers in columns
It has missing data. Markers are scored as -1 or 1 (no heterozygot scores)
`mrkDatNoNA` is a matrix. Same scoring but all missing marker data imputed
```{r read marker data}
mrkDat <- readr::read_tsv(
  file=here::here("data", "BreedBaseGenotypesDownloadT3.tsv"))
# Check what the different allele calls are
mrkDat[,-1] %>% unlist %>% unique
# Simple filtering to remove accessions and markers with too much missing data
nIsNAbyMrk <- apply(mrkDat, 1, function(v) sum(is.na(v)))
# Start by removing all markers that are all missing: that's just confusing
mrkDat <- mrkDat[nIsNAbyMrk < 216,]
# Remove markers and accessions with > 20% missing scores
nIsNAbyMrk <- apply(mrkDat, 1, function(v) sum(is.na(v)))
mrkDat <- mrkDat[nIsNAbyMrk / ncol(mrkDat) < 0.2,]
nIsNAbyAcc <- apply(mrkDat, 2, function(v) sum(is.na(v)))
mrkDat <- mrkDat[,nIsNAbyAcc / nrow(mrkDat) < 0.2]

# I like having accessions in rows and markers in columns, so change it
mrkDatT <- mrkDat %>% pivot_longer(cols= -1) %>% 
  pivot_wider(names_from = "Marker", values_from = "value") %>% 
  rename(accession = name)
# I like having marker scores be -1, 0, 1
mrkDatT[,-1] <- mrkDatT[,-1] - 1
mrkDatNoNA <- impute.glmnet(mrkDatT[,-1])
rownames(mrkDatNoNA) <- mrkDatT$accession
```

## Read and wrangle the phenotypic data  
After this, `phenDat` is a tibble with columns for yield, test weight, and
plant height
```{r read phenotypic data}
phenDat <- readr::read_csv(
  file=here::here("data", "BreedBaseMultitrialPhenoDownloadT3.csv"), skip=3)
phenDat <- phenDat %>% 
  rename(yield = `Grain yield - kg/ha|CO_321:0001218`,
         testWeight = `Grain test weight - g/l|CO_321:0001210`,
         height = `Plant height - cm|CO_321:0001301`) %>% 
  select(locationName, studyName, studyYear, germplasmName,
         replicate, blockNumber, plotNumber,
         rowNumber, colNumber,
         yield, testWeight, height) %>%
  arrange(locationName, replicate, plotNumber)
```

